The fundamental shape of dynamic experiences is preserved in episodic memory

### 2018-07-09 main.tex
checking in temporal and semantic clustering methods

\subsubsection{The fundamental shape of an experience is preserved in recall}
The analyses described in the previous section are useful in that they allow us to quantify memory dynamics for naturalistic stimuli in a way that is comparable to a vast literature on list-learning. However, these methods (and trial-based experiments more generally) fail to capture the rich semantic and temporal structure present in naturalistic stimuli and associated recall of those experiences. Thus, our next set of analyses test whether the `shape' of an experience is preserved in memory. We hypothesized that despite substantial individual variability in the precision and amount of content recalled across by participants, the overall shape of the episode would be recapitulated by all.

@@

\textbf{Serial position curve (SPC).} The SPC represents the proportion of participants that remember an item as a function of its serial position during encoding. We initialized a \# of participants (17) by \# of video events (34) matrix. Then, for each recall event (and each participant), we found the index of the video event that was recalled and filled it in with a 1. This resulted in a matrix where 1s indicate the successful recall of an event in serial position $n$ and zeros indicate the lack of recall for that event.  Lastly, we averaged over the rows of the matrix to get a 1 by 34 array representing the proportion of subjects that recalled an event as a function of its serial position.
% We used a few different metrics to assess memory that were all designed to capture the relationship between the video and recall models.  First, to get a general sense for the match between the video and recall matrices, we vectorized the the matrices and correlated them.  This resulted in a single number representing how well a given participant recalled the video. To get a temporally dynamic measure of memory, for each participant we computed the pairwise correlation between matching video/recall timepoints.  This allowed us to assess memory for individual moments of the video. Finally to get a more nuanced representation of memory for the video, we computed a video/recall model correlation matrix for each participant. These timepoint (1976) by timepoint (1976) matrices represent the correlation between every moment of the video model and every moment of the recall model. To provide some intuition, if a participant recalled every scene at the same rate and in exactly the same order as the video, we would expect high correlation values along the diagonal of this matrix. However, if a participant recalls the scenes out of order, or at different rates for different scenes, this will result in high values off the diagonal.
%
% The only additional constraint we imposed was that there was at least 2 points between each line segment.
% We searched for the optimal number of line segments by iteratively splitting the participants into two groups, fitting the model to the first half and testing the model on the remaining subjects. To compute fit to the model, for each recall event (a 1 by 100 topics vector) we computed the nearest line segment and then calculated the mean squared distance averaging over all events and all participants (in the sample). We repeated this 20 times using 1-10
%
% \subsubsection{Bridging between the stimulus and recall}
% The participants in this study were instructed to recall as much of the episode as they could in the order it happened.  Thus, we hypothesized that the recall models should resemble the video model to the extent that participants followed these instructions and had reasonably good memory accuracy. To test whether this was true, we correlated each moment of the video model with each moment of each participants' resampled recall model. We then averaged these matrices together to get a single group averaged matrix.  The result was a viewing time (1976) by recall time (1976) correlation matrix representing the average relationship between every moment of the movie and recall models (Fig.~\ref{fig:model}e). Notably, we observe significant correlation values (p<.05, permutation test described in \ref{sec:methods}) primarily along the diagonal, suggesting that our model captures the fact that participants recalled the episode in order. We also visualized this effect by projecting the matrices into a 3D space (reduced using Spectral Embedding [cite]) where the proximity of the trajectories at every timepoint represents how similar the topic vectors were. Visualizing the data in this way makes it clear that the overall shape and correlational structure of the video and average recall matrices is quite similar.

>>>>

\subsubsection{The fundamental shape of an experience is preserved in recall}
The analyses described in the previous section are useful in that they allow us to quantify memory dynamics for naturalistic stimuli in a way that is comparable to a vast literature on trial-based list-learning experiments. However, these methods fail to capture the rich temporal structure present in naturalistic stimuli and associated recall of those experiences. Thus, our next set of analyses test whether the ``shape'' of an experience is preserved in memory. We hypothesized that despite substantial individual variability in the precision and amount of content recalled across by participants, the overall shape of the episode would be recapitulated by the majority of participants.

@@

\textbf{Serial position curve (SPC).} The SPC represents the proportion of participants that remember an item as a function of its serial position during encoding. We initialized a \# of participants (17) by \# of video events (34) matrix. Then, for each recall event (and each participant), we found the index of the video event that was recalled and filled it in with a 1. This resulted in a matrix where 1s indicate the successful recall of an event in serial position $n$ and zeros indicate the lack of recall for that event.  Lastly, we averaged over the rows of the matrix to get a 1 by 34 array representing the proportion of subjects that recalled an event as a function of its serial position.

\textbf{Temporal clustering.} Temporal clustering measures the extent to which participants group their recall responses according to encoding position. For instance, if the participant recalled each item in the presentation order, this would result in a score of 1. If the participant recalled randomly with respect to presentation order, this would result in a score of ~.5.  For each event transition (and separately for each participant), we computed the rank similarity (euclidean distance) between the presentation position  of the current and next recall events. The scores were then averaged within participant to get a single number representing the extent of temporal clustering exhibited by a given participant.

\textbf{Semantic clustering.} Similar to temporal clustering, semantic clustering measures the extent to which participants group their recall responses according to semantic similarity. Here, we are using the topic vectors for each event as a proxy for its semantic content. Thus, similarity between the semantic content for two events can be computed by correlating their respective topic vectors.  For instance, if each consecutive recall was the next most similar event (in terms of its s), this would result in a score of ~1. If the participant recalled randomly with respect to semantic similarity, this would result in a score of ~.5.  For each event transition (and separately for each participant), we computed the rank similarity (correlation distance) between the current recall event and the next recall event. The scores were then averaged within participant to get a single number representing the extent of semantic clustering exhibited by a given participant.

% We used a few different metrics to assess memory that were all designed to capture the relationship between the video and recall models.  First, to get a general sense for the match between the video and recall matrices, we vectorized the the matrices and correlated them.  This resulted in a single number representing how well a given participant recalled the video. To get a temporally dynamic measure of memory, for each participant we computed the pairwise correlation between matching video/recall timepoints.  This allowed us to assess memory for individual moments of the video. Finally to get a more nuanced representation of memory for the video, we computed a video/recall model correlation matrix for each participant. These timepoint (1976) by timepoint (1976) matrices represent the correlation between every moment of the video model and every moment of the recall model. To provide some intuition, if a participant recalled every scene at the same rate and in exactly the same order as the video, we would expect high correlation values along the diagonal of this matrix. However, if a participant recalls the scenes out of order, or at different rates for different scenes, this will result in high values off the diagonal.
%
% The only additional constraint we imposed was that there was at least 2 points between each line segment.
% We searched for the optimal number of line segments by iteratively splitting the participants into two groups, fitting the model to the first half and testing the model on the remaining subjects. To compute fit to the model, for each recall event (a 1 by 100 topics vector) we computed the nearest line segment and then calculated the mean squared distance averaging over all events and all participants (in the sample). We repeated this 20 times using 1-10
%
% \subsubsection{Bridging between the stimulus and recall}
% The participants in this study were instructed to recall as much of the episode as they could in the order it happened.  Thus, we hypothesized that the recall models should resemble the video model to the extent that participants followed these instructions and had reasonably good memory accuracy. To test whether this was true, we correlated each moment of the video model with each moment of each participants' resampled recall model. We then averaged these matrices together to get a single group averaged matrix.  The result was a viewing time (1976) by recall time (1976) correlation matrix representing the average relationship between every moment of the movie and recall models (Fig.~\ref{fig:model}e). Notably, we observe significant correlation values (p<.05, permutation test described in \ref{sec:methods}) primarily along the diagonal, suggesting that our model captures the fact that participants recalled the episode in order. We also visualized this effect by projecting the matrices into a 3D space (reduced using Spectral Embedding [cite]) where the proximity of the trajectories at every timepoint represents how similar the topic vectors were. Visualizing the data in this way makes it clear that the overall shape and correlational structure of the video and average recall matrices is quite similar.
